\documentclass[12pt, letterpaper]{article}

\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{hyperref}
\hypersetup{
	%pagebackref=true,
	pdftitle={\@title},
	pdfauthor={\@author},
	pdfsubject={Essay about Bitcoins, with facts and questions},
	pdfcreator={LaTeX with abnTeX2},
	pdfkeywords={bitcoins criptocurrencies digital currencies},
	colorlinks=true,            % false: boxed links; true: colored links
	linkcolor=blue,             % color of internal links
	citecolor=blue,             % color of links to bibliography
	filecolor=magenta,          % color of file links
	urlcolor=blue,
	bookmarksdepth=4
}



\title{The digital era of information overflow and its underlying ``choice paradox''}
\author{Bruno Saboia de Albuquerque\thanks{Many thanks for my co-workers at the Bank for International Settlements (\url{https://bis.org}) for their immeasurably valuable help.}\and Christophe Laforge}

\date{December 2022}

\begin{document}
	\maketitle

	\begin{abstract}
		The classic microeconomics approach is to consider that more choices are always better for the customer -- a concept that has been conveyed before by folklore, with the well-known saying: ``the more, the merrier''. However, it might be the case that more choice is actually harmful for the consumer. Our goal is to investigate when that happens, and devise optimal strategies for the consumer to choose given an overwhelming amount of choices -- common on the current era of streaming platforms and digital information.

		%\vspace{\onelineskip}
		\noindent
		\\\textbf{Key-words}: Microeconomics. Consumer choice. Choice paradox.

	\end{abstract}
	\clearpage

	\section{Introduction}
	The modeling of consumer's choice within a mathematical framework certainly helped economists to have a better understanding of how consumers behave in reality. Albeit limited, these models bring notions that are useful for students, teachers, researchers and those who are interested in predicting how a consumer will behave in the future, or explain why it behaved in a certain way in the past.

	One of such notion is that of \textbf{preferences}. The literature regarding this subject is immense. Jehle and Reny \cite{jehle2011advanced}, for example, state that ``preferences are characterized axiomatically'', and afterwards define the two main axioms of consumer preferences: \textit{completeness} and \textit{transitivity}. Those are in the core of microeconomics theory of choice.
	
	Preferences are stated in terms of \textbf{utility}. If a given good yields more utility than some other, the more ``useful'' one is preferred over the other. Although this seems fairly simple and intuitive, it is difficult to measure utility. 
	
	Naturally, the perceived utility will depend on the situation and the consumer. Different consumers, facing different situations, will assign different utilities for the same good, commodity or service. Therefore the assignment and thus ordering of utility between different goods is a subjective matter, always subject to change. Even when subject to the same budget constraint, the same consumer will change their preference, given a different situation.
	
	Intuitively, it seems better to have more choices than fewer. On developed economies, consumers can enjoy a multitude of options, thus their can order their preferences in a way that is more satisfactory to them. This is seemingly better than in underdeveloped or authoritarian countries, where choices can be limited. In the end, consumers, no matter how wealthy they are, cannot improve their life quality without choices. 
	
	However, the era of internet overwhelmed consumers with different options, bringing an apparent paradox: it seems that having \textit{too much} choice is, in the end, harmful. Some online services, such as movie streaming, provide so many choices that consumers cannot choice which movie to watch -- and give up of using the service.
	
	Our goal in this article is to understand whether is the case that more choices can be harmful -- also trying to understand when in fact many options start to jeopardize utility. We are also interested in devising an optimal choice strategy for a consumer in the current world of many choices. 


	\section{Preferences axioms}
	In the introduction, we said that consumer preferences are usually characterized by an axiomatic system. The two main axioms are \textit{completeness} and \textit{transitivity}.  
	
	Explaining the axioms in laymen terms, we can say that \textit{completeness means that consumers are able to evaluate their preferences between different given alternatives}. That is, given a finite number of choices, the consumer will be able to rank them, from best to worse. \textit{Transitivity means that the consumer choice is consistent}, that is, if choice A is preferred to B, and B is preferred to C, then A is also preferred to C. Preferences are stated in terms of utility. Therefore, a choice A is said preferred to B if A has at least the same utility as B. The usual mathematical notation for representing this statement is $A \succsim B$\footnote{For our purposes, it is unnecessary to make a distinction between preference and strict preference relations.}.

	Going further and explaining transitivity in the usual binary relation notation, let $X$ be the consumption set, and let $x_i \in \mathbb{R}$ be the number of units of commodity $i$\footnote{Hence, the commodity quantity is expressed in some infinitely divisible units.}, and let $x^k= (x_1, x_2, \dots, x_n) \in X$ be the consumption bundle $k$, being $n$ a finite number. Assume that $X \subseteq \mathbb{R}_+^n$, $0 \in X$, and furthermore that $X$ is closed and convex. Let $x^1, x^2, x^3 \in X$ be consumption bundles. If the following two statements hold:

	\begin{equation}
		x^1 \succsim x^2; x^2 \succsim x^3
	\end{equation}

	Then the following statement will also necessarily hold:

	\begin{equation}
		x^1 \succsim x^3
	\end{equation}

	The transitivity axiom might seem fairly intuitive. If one prefers apples to oranges, and oranges to bananas, intuition can tell that one will also prefer apples to bananas. However, empirical evidence seems to demonstrate that preferences are in reality intransitive \cite{GUADALUPELANAS2020e03459}. 
	
	This perhaps raise suspicion on the completeness axiom, which by some naïve intuition can already feel wrong: to be able to completely evaluate one's preferences, one must be able to know the utility of the alternative. In many cases, consumer's do not necessarily know if a different choice would bring a better utility. Therefore, real-life choice seems to involve uncertainty -- something not captured by this simple model of consumer choice.

	\section{Utility and choices}
	Usually, the bundles referred in the preferences theory are composed of multiple elements. Naturally, a bundle can consist of a single element -- for that, the vector of quantities is $0$ for all commodities, except from one. That is, let  $y > 0$ and let $x = (y_1, 0_{2}, 0_{3}, \dots, 0_n)$ be a bundle with $n$ items. Only the first item has any quantity on it. Therefore, this bundle can be simplified to $x = (y_1)$.  
		
	In our case, for the sake of simplification, let us assume that our bundles contain a single element as described above, and that for each bundle, $x^k = (x_i), x^{k+1}= (x_{i+1}), \dots , x^{k+j} = (x_{i+j})$ -- that is, no bundle contains an element already present in another bundle. In other words, all bundles consists of a single element, which is unique across the bundle set. Let us call such bundle set a \textit{choice}. Let also $k+j$ be finite, so there is a finite number of choices. The sum $k+j$ is therefore the total number of choices.
	
	Let $u(x^k) \in \mathbb{R}_+^n$ be the utility function of the $k$-th choice, and let $U_k$ be the sum of the utilities of a given set of choices $(x^1, \dots , x^k)$, so that:
	 
	\[ U_k = \sum_{n=1}^{k} u(x^n) \]
	
	We cannot infer from the axioms that the following does not hold:  
	
	\begin{equation}
		U_k \leq U_{k+j} \forall j \in \mathbb{N}_{> 0}
	\end{equation}

	That is, the total utility yielded by a set of choices with $k$ elements is at least as good as a set with $k+j$ elements. Therefore, we can assume that \textit{utility is non-decreasing on choice}. One is never worse of, in terms of utility, by having an additional choice. 

	Although Pareto observed that utility is incomparable between two different consumers and that a measurable value for the utility function was not necessary for the demand theory \cite{pareto1897}, the notion of a negative value for the utility function of a given consumption bundle seems indeed counter intuitive. Of course, the actual consumption of some goods -- such as alcoholic spirits or cigarettes -- may be harmful to the customer. However, let us ignore this possibility, and do a thought experiment. 
	
	Imagine a customer with no budget restriction. Additionally, imagine that such customer does not infer in storage costs -- or any other cost for holding an additional choice. Such customer likes bananas and apples. Now, suppose that he is presented with two choices: one banana, or one banana and one apple. It is logical that the first set, containing only one banana, yields \textit{at least} the same utility than one banana and one apple.
	
	Therefore, it seems counter-intuitive that it is the number of available choices itself that brings harm to the consumer.
	
	\subsection{Ranking choices}
	Recall that, by definition, the number of items in a given consumption bundle is finite. This condition is necessary so that ranking between different bundle sets is possible. If there were infinite commodities in a given consumption bundle, then its utility would be ill-defined, thus ranking a bundle containing infinite elements against a different bundle would not be possible.
	
	In the simplified consumer model, the ranking of choices is already given.


\bibliographystyle{unsrt}
\bibliography{decision-paradox-paper}
\end{document}
